\begin{abstract}
  This thesis reports on the development of a family of parallel algorithms for
  computing area openings and closings on gray-scale images, in order to
  increase performance on large images. Previous parallel and sequential
  algorithms did not fully exploit the parallel nature of contemporary
  computers. The novelty in this thesis is that the developed algorithms use
  variants of a shared union-find data structure. This shared union-find
  approach reduces the amount of sequential work substantially.

  By conducting structured performance experiments, I show that, in particular,
  a certain type of the new, parallel algorithms outperforms the original,
  sequential algorithm by far on large images. Moreover, by identifying formal
  correctness properties and by using Java Pathfinder to model check the
  respective implementations, I show that the new parallel algorithms are
  formally correct. Additionally, I discuss the feasibility of model checking
  complex image filtering algorithms.

  Throughout the thesis, I review different variants of concurrent union-find
  and asses their performance on random graphs, as well as their adaption to
  parallel area opening. The presented results show that the structure of random
  input graphs has a greater influence on performance than their mere
  size. Furthermore, I discuss input image structure in relation to the
  performance, exhibited by the various parallel area opening
  algorithms. Performance experiments show that performance limitations on
  random graphs do not necessarily apply to the performance on images. This
  thesis opens up for many future research directions on parallel morphological
  algorithms.
\end{abstract}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
